---
title: "Research Project: TMLE Implementation"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "Sylvia Cheng"
output: 
  pdf_document:
    latex_engine: xelatex
---
# set up 

```{r library, echo = TRUE, error = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE, 
                      error = FALSE, warning = FALSE, message = FALSE)

library(data.table)
library(janitor)
library(tidyverse)
library(flextable)
library(sl3)
library(tmle)
library(tmle3)
library(SuperLearner)
library(origami)
library(ranger)
```

# prepare data

```{r, data-read-in}
data <- read_csv("data_tmle_analysis.csv") %>% 
  mutate(hv_1216 = hv_1216/1000, # make the unit of all $ variables per 1000
         hv_1519 = hv_1519/1000,
         income_1216 = income_1216/1000,
         totrev_1617 = totrev_1617/1000,
         totrev_1415 = totrev_1415/1000,
         totrev_higher_than_med = case_when(totrev_1617 - totrev_1617 %>% median() >= 0 ~ 1, TRUE ~ 0), # treatment
         hv_higher_than_med = case_when(hv_1216 - hv_1216 %>% median() >= 0 ~ 1, TRUE ~ 0), # stratification 1
         percblack_higher_than_med = case_when(perc_black_1314 - perc_black_1314 %>% median() >= 0 ~ 1, TRUE ~ 0)
         ) %>% # stratification 2
  filter(!District == "Los Angeles Unified") 

data.table(data) -> data_tabled

colnames(data_tabled) # show all the variables we have
```


# make SL libraries

```{r}
lrnr_glm <- Lrnr_glm$new()
lrnr_gam <- Lrnr_gam$new()
lrnr_lasso <- Lrnr_glmnet$new(alpha = 1) 
lrnr_enet <- Lrnr_glmnet$new(alpha = 0.5) 
lrnr_ridge <- Lrnr_glmnet$new(alpha = 0) 
lrnr_polspline <- Lrnr_polspline$new() # can pick up interactions
lrnr_earth <- Lrnr_earth$new()
lrnr_nnls <- Lrnr_nnls$new()
lrnr_ranger <- Lrnr_ranger$new() # also picks up inter through the procedure, fast!
lrnr_xgboost100 <- Lrnr_xgboost$new(nrounds = 100, early_stopping_rounds = 10) # SAME HERE, any treee based learners 

discrete <- Lrnr_cv_selector$new(loss_squared_error) # our meta-learner

sl_g <- Lrnr_sl$new(
  list(lrnr_glm, lrnr_earth, lrnr_ridge, lrnr_ranger),
  discrete
)

# build interaction lrnr
vars_forlist_black <- c("dist_size_1314", "totrev_1415", "gp_perc_1216", "pov_1216", "hv_1216")
MyList_black <- list()
for (var in 1:length(vars_forlist_black)){
  MyList_black[[var]] <- c("totrev_higher_than_med", vars_forlist_black[var])
} 
lrnr_interact_black <- Lrnr_define_interactions$new(interactions = MyList_black)
lrnr_glm_interact_full_black <- make_learner(Pipeline, lrnr_interact_black, lrnr_glm)

vars_forlist_hv <- c("dist_size_1314", "perc_black_1314", "totrev_1415", "gp_perc_1216", "pov_1216")
MyList_hv <- list()
for (var in 1:length(vars_forlist_hv)){
  MyList_hv[[var]] <- c("totrev_higher_than_med", vars_forlist_hv[var])
} 
lrnr_interact_hv <- Lrnr_define_interactions$new(interactions = MyList_hv)
lrnr_glm_interact_full_hv <- make_learner(Pipeline, lrnr_interact_hv, lrnr_glm)


# learners for each interaction term 
term_dist1314 <- list(c("totrev_higher_than_med", "dist_size_1314"))
term_percblack1314 <- list(c("totrev_higher_than_med", "perc_black_1314"))
term_totrev1415 <- list(c("totrev_higher_than_med", "totrev_1415"))
term_gpperc1216 <- list(c("totrev_higher_than_med", "gp_perc_1216"))
term_pov1216 <- list(c("totrev_higher_than_med", "pov_1216"))
term_hv1216 <- list(c("totrev_higher_than_med", "hv_1216"))

lrnr_interact_term_dist1314 <- Lrnr_define_interactions$new(interactions = term_dist1314)
lrnr_glm_interact_term_dist1314 <- make_learner(Pipeline, lrnr_interact_term_dist1314, lrnr_glm)

lrnr_interact_term_percblack1314 <- Lrnr_define_interactions$new(interactions = term_percblack1314)
lrnr_glm_interact_term_percblack1314 <- make_learner(Pipeline, lrnr_interact_term_percblack1314, lrnr_glm)

lrnr_interact_term_totrev1415 <- Lrnr_define_interactions$new(interactions = term_totrev1415)
lrnr_glm_interact_term_totrev1415 <- make_learner(Pipeline, lrnr_interact_term_totrev1415, lrnr_glm)

lrnr_interact_term_gpperc1216 <- Lrnr_define_interactions$new(interactions = term_gpperc1216)
lrnr_glm_interact_term_gpperc1216 <- make_learner(Pipeline, lrnr_interact_term_gpperc1216, lrnr_glm)

lrnr_interact_term_pov1216 <- Lrnr_define_interactions$new(interactions = term_pov1216)
lrnr_glm_interact_term_pov1216 <- make_learner(Pipeline, lrnr_interact_term_pov1216, lrnr_glm)

lrnr_interact_term_hv1216 <- Lrnr_define_interactions$new(interactions = term_hv1216)
lrnr_glm_interact_term_hv1216 <- make_learner(Pipeline, lrnr_interact_term_hv1216, lrnr_glm)

# different SL of Q for different stratifications ----
sl_Q_black <- Lrnr_sl$new(
  list(lrnr_glm, lrnr_earth, lrnr_lasso, lrnr_enet, lrnr_ridge, lrnr_ranger, 
       lrnr_glm_interact_full_black, lrnr_glm_interact_term_hv1216,
       lrnr_glm_interact_term_dist1314, lrnr_glm_interact_term_totrev1415, 
       lrnr_glm_interact_term_gpperc1216, lrnr_glm_interact_term_pov1216
  ), 
  discrete
)

sl_Q_hv <- Lrnr_sl$new(
  list(lrnr_glm, lrnr_earth, lrnr_lasso, lrnr_enet, lrnr_ridge, lrnr_ranger, 
       lrnr_glm_interact_full_hv, lrnr_glm_interact_term_dist1314, 
       lrnr_glm_interact_term_percblack1314, lrnr_glm_interact_term_totrev1415, 
       lrnr_glm_interact_term_gpperc1216, lrnr_glm_interact_term_pov1216
  ), 
  discrete
)

learner_list_black <- list(Y = sl_Q_black, A = sl_g)
learner_list_hv <- list(Y = sl_Q_hv, A = sl_g)

W_black <-c("dist_size_1314", "totrev_1415", "gp_perc_1216", "pov_1216", "income_1216", "hv_1216")
W_hv <-c("dist_size_1314", "perc_black_1314", "totrev_1415", "gp_perc_1216", "pov_1216", "income_1216")

node_list_black <- list(W = W_black, V = "percblack_higher_than_med", A = "totrev_higher_than_med", Y = "perc_black_1718")
node_list_hv <- list(W = W_hv, V = "hv_higher_than_med", A = "totrev_higher_than_med", Y = "perc_black_1718")

ate_spec <- tmle_ATE(treatment_level = 1, control_level = 0) 
spec <- tmle_stratified(ate_spec)
```


# stratified TMLE: housing value

```{r}
set.seed(5824)
stratified_fit_hv <- tmle3(spec, data_tabled, node_list_hv, learner_list_hv)
stratified_fit_hv
```

# stratified TMLE: black student enrollment

```{r}
set.seed(5824)
stratified_fit_black <- tmle3(spec, data_tabled, node_list_black, learner_list_black)
stratified_fit_black

```

```{r}
# result_param <- c(param = stratified_fit_black$summary$param)
# result_ATE <- c(ATE = stratified_fit_black$summary$tmle_est)
# result_SE <- c(SE = stratified_fit_black$summary$se)
# result_CI_lower <- c(CI_lower = stratified_fit_black$summary$lower)
# result_CI_upper <- c(CI_upper = stratified_fit_black$summary$upper) 
# 
# result_param %>% as_tibble() %>% rename(param = value) -> result_param
# result_ATE %>% as_tibble() %>% rename(ATE = value) -> result_ATE
# result_SE %>% as_tibble() %>% rename(SE = value) -> result_SE
# result_CI_lower %>% as_tibble() %>% rename(CI_lower = value) -> result_CI_lower
# result_CI_upper %>% as_tibble() %>% rename(CI_upper = value) -> result_CI_upper
# 
# 
# result_param %>% bind_cols(result_ATE) %>% bind_cols(result_SE) %>% bind_cols(result_CI_lower) %>%
#   bind_cols(result_CI_upper) %>%
#   mutate(ATE = (exp(ATE)-1)*100,
#          SE = (exp(SE)-1)*100, 
#          CI_lower = (exp(CI_lower)-1)*100,
#          CI_upper = (exp(CI_upper)-1)*100
#          ) %>% 
#   flextable() %>% autofit() %>% add_header_lines("table. exponentiated results of stratification on black student enrollment")

```


# TMLE: no stratification

```{r}
vars_forlist_nostrat <- c("dist_size_1314", "totrev_1415", "gp_perc_1216", "pov_1216", "hv_1216", "perc_black_1314")
MyList_nostrat <- list()
for (var in 1:length(vars_forlist_nostrat)){
  MyList_nostrat[[var]] <- c("totrev_higher_than_med", vars_forlist_nostrat[var])
} 
lrnr_interact_nostrat <- Lrnr_define_interactions$new(interactions = MyList_nostrat)
lrnr_glm_interact_full_nostrat <- make_learner(Pipeline, lrnr_interact_nostrat, lrnr_glm)

sl_Q_nostrat <- Lrnr_sl$new(
  list(lrnr_glm, lrnr_earth, lrnr_lasso, lrnr_enet, lrnr_ridge, lrnr_ranger, 
       lrnr_glm_interact_full_nostrat, lrnr_glm_interact_term_hv1216,
       lrnr_glm_interact_term_dist1314, lrnr_glm_interact_term_totrev1415, 
       lrnr_glm_interact_term_gpperc1216, lrnr_glm_interact_term_pov1216, lrnr_glm_interact_term_percblack1314
  ), 
  discrete
)

learner_list_nostrat <- list(Y = sl_Q_nostrat, A = sl_g)

W_nostrat <-c("dist_size_1314", "perc_black_1314", "totrev_1415", "gp_perc_1216", "hv_1216", "pov_1216", "income_1216")

node_list_nostrat <- list(W = W_nostrat, A = "totrev_higher_than_med", Y = "perc_black_1718")

set.seed(5824)
nostrat_fit <- tmle3(ate_spec, data_tabled, node_list_nostrat, learner_list_nostrat)
nostrat_fit
```


# CV risks, propensity scores, and variable importance

## functions 

```{r}
# function to get SL's CV risk table for predicting **A** from tmle3 fit
get_cvrisk_tbl_A <- function(tmle3_fit){
  cvrisk_tbl_A <- tmle3_fit$likelihood$initial_likelihood$factor_list$A$learner$cv_risk(loss_squared_error)[,c(1,3)]
  cvrisk_tbl_A$coefficients <- c(tmle3_fit$likelihood$initial_likelihood$factor_list$A$learner$coefficients, NA)
  return(cvrisk_tbl_A)
}

# function to get SL's CV risk table for predicting **Y** from tmle3 fit
get_cvrisk_tbl_Y <- function(tmle3_fit){
  cvrisk_tbl_Y <- tmle3_fit$likelihood$initial_likelihood$factor_list$Y$learner$cv_risk(loss_squared_error)[,c(1,3)]
  cvrisk_tbl_Y$coefficients <- c(tmle3_fit$likelihood$initial_likelihood$factor_list$Y$learner$coefficients, NA)
  return(cvrisk_tbl_Y)
}

# function to get plot of propensity scores (PS) to examine positivity
get_PSplot <- function(tmle3_fit){
  propensity_score <- tmle3_fit$likelihood$initial_likelihood$factor_list$A$learner$predict()
  A_factor <- ifelse(data_tabled$totrev_higher_than_med == 1, "Group with A=1", 
                     "Group with A=0")
  plot_df <- data.frame(Group = as.factor(A_factor), PS = propensity_score)
  PSplot <- ggplot(plot_df, aes(x = PS, color = Group)) + 
    geom_histogram(fill = "white",  position = "dodge", binwidth = .01) +
    geom_density(alpha = .2) +
    xlim(c(0,1)) +
    theme(legend.title = element_blank()) +
    scale_color_brewer(palette="Accent") + 
    theme_minimal() +
    theme(legend.position = "bottom") +
    labs(title="Distribution of Propensity Scores", x="Propensity Score", y="Counts")
  return(PSplot)
}

```

## stratified: black student enrollment

### A

```{r}
get_cvrisk_tbl_A(stratified_fit_black)
```

### Y

```{r}
get_cvrisk_tbl_Y(stratified_fit_black)
```

### propensity score plot 

```{r}
get_PSplot(stratified_fit_black)
```

### by the median of black student enrollment

```{r}
get_PSplot_black <- function(tmle3_fit){
  propensity_score <- tmle3_fit$likelihood$initial_likelihood$factor_list$A$learner$predict()
  A_factor <- ifelse(data_tabled$totrev_higher_than_med == 1, "Group with A=1", 
                     "Group with A=0")
  black_yn <- ifelse(data_tabled$percblack_higher_than_med == 1, "black student enrollment >= median", 
                     "black student enrollment < median")
  plot_df_black <- data.frame(Group = as.factor(A_factor), PS = propensity_score, stratifier = as.factor(black_yn))
  PSplot_black <- ggplot(plot_df_black, aes(x = PS, color = Group)) + 
    geom_histogram(fill = "white",  position = "dodge", binwidth = .01) +
    facet_wrap(vars(stratifier)) +
    geom_density(alpha = .2) +
    xlim(c(0,1)) +
    theme(legend.title = element_blank()) +
    scale_color_brewer(palette="Accent") + 
    theme_minimal() +
    theme(legend.position = "bottom") +
    labs(title="Distribution of Propensity Scores", x="Propensity Score", y="Counts")
  return(PSplot_black)
}

get_PSplot_black(stratified_fit_black)
```

## stratified: housing value 

### A

```{r}
get_cvrisk_tbl_A(stratified_fit_hv)
```

### Y

```{r}
get_cvrisk_tbl_Y(stratified_fit_hv)
```

### propensity score plot 

```{r}
get_PSplot(stratified_fit_hv)
```

### by the median of housing value

```{r}
get_PSplot_hv <- function(tmle3_fit){
  propensity_score <- tmle3_fit$likelihood$initial_likelihood$factor_list$A$learner$predict()
  A_factor <- ifelse(data_tabled$totrev_higher_than_med == 1, "Group with A=1", 
                     "Group with A=0")
  hv_yn <- ifelse(data_tabled$hv_higher_than_med == 1, "housing value >= median", 
                     "housing value < median")
  plot_df_hv <- data.frame(Group = as.factor(A_factor), PS = propensity_score, stratifier = as.factor(hv_yn))
  PSplot_hv <- ggplot(plot_df_hv, aes(x = PS, color = Group)) + 
    geom_histogram(fill = "white",  position = "dodge", binwidth = .01) +
    facet_wrap(vars(stratifier)) +
    geom_density(alpha = .2) +
    xlim(c(0,1)) +
    theme(legend.title = element_blank()) +
    scale_color_brewer(palette="Accent") + 
    theme_minimal() +
    theme(legend.position = "bottom") +
    labs(title="Distribution of Propensity Scores", x="Propensity Score", y="Counts")
  return(PSplot_hv)
}

get_PSplot_hv(stratified_fit_hv)
```


## no-stratification 

### A

```{r}
get_cvrisk_tbl_A(nostrat_fit)
```

### Y

```{r}
get_cvrisk_tbl_Y(nostrat_fit)
```

### propensity score plot 

```{r}
get_PSplot(nostrat_fit)
```


# non-TMLE stat for comparison

## t-test by the median of housing value at baseline

```{r}
t.test(totrev_higher_than_med ~ hv_higher_than_med, data = data_tabled, var.equal = FALSE) -> test

test$stderr
```

## t-test by the median of black student enrollment at baseline 

```{r}
t.test(totrev_higher_than_med ~ percblack_higher_than_med, data = data_tabled, var.equal = FALSE)
```

## glm model 

```{r}
lm.model <- glm(totrev_higher_than_med ~ dist_size_1314 + perc_black_1314 + totrev_1415 + gp_perc_1216 + pov_1216, data = data_tabled, family = binomial() )

summary(lm.model)

```
